Using tours to visually investigate properties of new projection pursuit indexes with application to problems in physics

### 2019-01-08 scagnostics.Rmd
Small fixes in Sec 3.6
Some comments added to Sec 4.2

Our results can be summarized by evaluating and comparing the advantages and disadvantages of each index function according to the criteria presented above. An overview is given in Table \ref{tab:summary}. The first three entries are from the scagnostics family. The advantages are good scaling of computing time with sample size and that the set of index functions covers general signatures, making the set flexible in applications (the same is however not true for each single index viewed on its own). On the other hand we have found that they are typically too noisy making optimization of the function challenging, which also results in poor squint ability. In particular the stringy index cannot be used with current optimizers. The index functions from @mbgraphic share similar behavior as they are both found to evolve fairly smoothly between interpolated projections, but have been shown to be strongly rotation dependent. The splines2D index is fast to calculate and found to be the most efficient method for identifying functional dependence. We have only considered the non-binned implementation of dcor2D which is slow to calculate on large datasets. While this index is in principle general, we have found it to be too sensitive to noise to be applicable in most realistic examples, therefore we consider flexibility limited. Finally the MINE indexes are found to be smooth only for large datasets, with some rotation dependence observed for the TIC index. While these indexes are slow to compute especially on large datasets, we have found them to be the most general, allowing good discrimination even with noisy structures. In addition we observed good squintability for all three examples considered, given large datasets. Results in Section \ref{sec:params} suggest that tuning of the MINE parameters may result in improved behavior for smaller datasets.

\begin{table}
\begin{center}
\caption{Summary of findings, showing to what extend the considered index functions pass the criteria for a good PPI. "\checkmark" symbols good behavior, "$\cdot$" symbols some issues while "$\times$" symbols failure on the corresponding criteria. Each index has particular strengths and drawbacks and selection must be guided by the considered example, see text for details.}
\label{tab:summary}
\begin{tabular}{|l|c|c|c|c|c|c|}
\hline
Index & smooth & squintability & flexible & rotation invariant & speed \\
\hline
Convex & $\cdot$ & $\cdot$ & $\cdot$ & $\cdot$ & \checkmark \\
Skinny & $\cdot$ & $\cdot$ & $\cdot$ & $\cdot$ & \checkmark \\
Stringy & $\times$ & $\times$ & $\cdot$ & $\cdot$ & \checkmark \\
\hline
splines2D & \checkmark & \checkmark & $\cdot$ & $\times$ & \checkmark \\
dcor2D & \checkmark & \checkmark & $\cdot$ & $\times$ & $\cdot$ \\
\hline
MINE & $\cdot$ & \checkmark  & \checkmark & $\cdot$ & $\cdot$ \\
\hline
\end{tabular}
\end{center}
\end{table}

@@

The first example shows 2538 posterior samples obtained by fitting source parameters to the observed gravitational wave signal GW170817 from a neutron star merger [@PhysRevLett119161101]. Data has been downloaded from @ligoData. The fitting procedures are described in detail in @Abbott2018exr. There are six parameters of physical interest (6D) with some known relationships. Projection pursuit is used to find the known relationships.

The second example shows data from @Smith:2016qas, generated from a simulation study of a binary black hole (BBH) merger event. (XXX More info needed.) There are 12 parameters (12-D), with multiple nuisance parameters. Projection pursuit uncovers several new relationships between parameters.

@@

Figure \ref{fig:bbhGuided3} shows six maxima obtained by different starts, for two types of parameters: first, spin related parameters (i.e. alpha, theta\_jn, chi\_tot and chi\_p), and second position related parameters (i.e. ra, dec and distance). Four of the six (a-d) are almost identical, but not interesting projections. Projection f has the highest PP index value but it is primarily the view seen in the bivariate plot of dec and ra.

Choosing a different subset of variables reveals something new. The subspace of m1, ra, chi\_tot, alpha, distance, dec produces a more refined view of Figure \ref{fig:bbhGuided3} projection f. When alpha contributes in contrast to dec, the relationship between the points is almost perfectly on a curve. This is shown in Figure \ref{constructedExample}. Manually reconstructing this can be done by differencing the two parameters, in their original units. \textcolor{red}{This result is unexpected, it can be understood from ????? (Does it mean that knowing the precision angle gives more precise information about the location? Knowing ra uniquely determines dec-alpha rather than dec it seems.)}

@@

>>>>

Our results can be summarized by evaluating and comparing the advantages and disadvantages of each index function according to the criteria presented above. An overview is given in Table \ref{tab:summary}. The first three entries are from the scagnostics family. The advantages are good scaling of computing time with sample size and that the set of index functions covers general signatures, making the set flexible in applications (the same is however not true for each single index viewed on its own). On the other hand we have found that they are typically too noisy making optimization of the function challenging, which also results in poor squint ability. In particular the stringy index cannot be used with current optimizers. The index functions from @mbgraphic share similar behavior as they are both found to evolve fairly smoothly between interpolated projections, but have been shown to be strongly rotation dependent. The splines2D index is fast to calculate and found to be the most efficient method for identifying functional dependence. We have only considered the non-binned implementation of dcor2D which is slow to calculate on large datasets. While this index is in principle general, we have found it to be too sensitive to noise to be applicable in most realistic examples, therefore we consider flexibility limited. Finally the MINE indexes are found to be smooth only for large datasets, with some rotation dependence observed for the TIC index. While these indexes are slow to compute especially on large datasets, we have found them to be the most general, allowing good discrimination even with noisy structures. In addition we observed good squintability for all three examples considered, given large datasets. Results in the appendix \ref{sec:params} suggest that tuning of the MINE parameters may result in improved behavior for smaller datasets.

\begin{table}
\begin{center}
\caption{Summary of findings, showing to what extend the considered index functions pass the criteria for a good PPI. "\checkmark" symbols good behavior, "$\cdot$" symbols some issues while "$\times$" symbols failure on the corresponding criteria. Each index has particular strengths and drawbacks and selection must be guided by the considered example, see text for details.}
\label{tab:summary}
\begin{tabular}{|l|c|c|c|c|c|c|}
\hline
Index & smooth & squintability & flexible & rotation invariant & speed \\
\hline
Convex & $\cdot$ & $\cdot$ & $\cdot$ & $\cdot$ & \checkmark \\
Skinny & $\cdot$ & $\cdot$ & $\cdot$ & $\times$ & \checkmark \\
Stringy & $\times$ & $\times$ & $\cdot$ & $\cdot$ & \checkmark \\
\hline
splines2D & \checkmark & \checkmark & $\cdot$ & $\times$ & \checkmark \\
dcor2D & \checkmark & \checkmark & $\cdot$ & $\times$ & $\cdot$ \\
\hline
MINE & $\cdot$ & \checkmark  & \checkmark & $\cdot$ & $\cdot$ \\
\hline
\end{tabular}
\end{center}
\end{table}

@@

The first example contains 2538 posterior samples obtained by fitting source parameters to the observed gravitational wave signal GW170817 from a neutron star merger [@PhysRevLett119161101]. Data has been downloaded from @ligoData. The fitting procedures are described in detail in @Abbott2018exr. There are six parameters of physical interest (6D) with some known relationships. Projection pursuit is used to find the known relationships.

The second example contains data generated from a simulation study of a binary black hole (BBH) merger event according to @Smith:2016qa. (XXX More info needed.) There are 12 parameters (12-D), with multiple nuisance parameters. Projection pursuit uncovers several new relationships between parameters.

@@

Figure \ref{fig:bbhGuided3} shows six maxima obtained by different starts, for two types of parameters: first, spin related parameters (i.e. alpha, theta\_jn, chi\_tot and chi\_p), and second position related parameters (i.e. ra, dec and distance). Four of the six (a-d) are almost identical, but not interesting projections. Projection f has the highest PP index value but it is primarily the view seen in the bivariate plot of dec and ra.  \textcolor{red}{Actually I don't think any of the obtained projections is particularly interesting on its own, but a good point to make here is that these are local maxima and moving between them should be revealing. E.g. I looked at a planned tour moving from views a-c-e-f which shows relation of multimodality in the high-D space, so maybe that is the point to make here?}

Choosing a different subset of variables reveals something new. The subspace of m1, ra, chi\_tot, alpha, distance, dec produces a more refined view of Figure \ref{fig:bbhGuided3} projection f. When alpha contributes in contrast to dec, the relationship between the points is almost perfectly on a curve. This is shown in Figure \ref{constructedExample}. Manually reconstructing this can be done by differencing the two parameters, in their original units. \textcolor{red}{This result is unexpected, it can be understood from ????? (Does it mean that knowing the precision angle gives more precise information about the location? Knowing ra uniquely determines dec-alpha rather than dec it seems.) Also highlighting why finding global maxima is important, ra, dec and alpha were present in the above example but the optimiser did not converge ot the the global maximum, thus the refined relationship was not revealed.}

@@

```{r eval=FALSE}
m3 <- matrix(c(1,0,0,0,1,0,0,0,1,0,0,0,1,0),ncol=2)
m4 <- matrix(c(1,1,1,0,0,0,0,0,1,1,0,0,0,0),ncol=2)
planeL <- list(m3, m4)
li <- 3
for(i in c(1, 3, 6, 5)){
  r <- resList[[i]]
  iLast <- length(r)
  fProj <- r[[iLast]]
  planeL[[li]] <- fProj
  li <- li+1
}
tP <- tourr::planned_tour(planeL[3:6]) # only need dummy planes when using history tour
dPoints <- attributes(r)$data
quartz()
animate_xy(dPoints, tP, pch=".")
```
